Introduction_outline

### 2019-01-22 Write_up/Writing/Introduction_outline.Rmd
minor_edits_refs

The duration and intensity of coastal upwelling are known to have a critical role in the phenology of key marine ecosystem processes. Similarly, changes in the characteristics of upwelling are shown to have cause large disturbances to ecosystems at several trophic levels [@wang]. Cold, nutrient rich, low pH, and low oxygen upwelled waters fuels primary production as a consequence of the development of phytoplankton blooms [@Ryther; @Cushing; @mcGillicuddy; @oschlies]. In these upwelling systems, the high productivity supports vast and diverse marine populations. Mesoscale and sub-mesoscale oceanic movements are key in managing the exchange and transportation of organic resources and marine organisms from the coastal zone to the open ocean. Most of the world’s major fishing industries are dependent on upwelling, because this process brings nutrient rich waters into the euphotic zone, where it forms the foundation of a localized food web that ecologically differs from surrounding, non-upwelled regions [@bosman; @barshis]. Whilst nutrients play a primary role in the survival of these systems, the cool temperatures of these waters are critical as well. There is an increasingly large body of research that has provided evidence showing that sea surface temperatures (SST) have and are continuing to increase. However, some regions still show a lower or non-significant increase, and in some cases cooling trends are evident (REFERENCES). Upwelling occurring along the Cape south coast is proven to have less of an economic importance in contrast to the west coast. Inshore regions and embayments that are protected from wind action may be subjected to intense episodes of primary production [@graham]. Vander Wouder et al. (2006) found that these sites represent the highest levels of chlorophyll concentration. Biotic systems within the Benguela upwelling system may be negatively influenced by the changes in oxygen levels given their adaption to natural variable conditions [@capone].

@@

Throughout region surrounding the Benguela EBUS, the atmosphere is dominated by anticyclonic high pressure cells with quasi-stationary positions. The South Atlantic Ocean High is situated along the west, drawing cool, dry air onto the west of the subcontinent [@van]. Solar heating during summer may result in the development of low pressure cells, known as heat lows. These cells are absent during winter time (Tyson and Preston-Whyte, 2000). When the anticyclones shift north, during winter months, the cold westerlies have a substantial impact on the weather of the southern tip of the South African subcontinent [@van]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van]. However, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings].

@@

Remotely-sensed satellite SST products are produced and maintained by a number of organisations worldwide. In order to obtain these SST data, advanced technology is launched into space, in order to obtain infra-red light being emitted by the ocean. This irradiance is then taken as a proxy for SST.  Due to the distance at which these SST are being collected, certain coarseness within the datasets are unavoidable. Satellite SST products undergo a range of quality control processes that are well researched and tested (Reynolds and Smith, 1994; Brown et al., 1999; Martin et al., 2012), making their use more reliable. However, Smit et al. (2013) have shown that SST data are not particularly reliable for coastal regions and should be used cautiously as they can be prone to both warm and cold bias. Nevertheless, satellite data products are extremely useful when investigating the ocean at broad scale and can provide important SST data.  <will add more here on the different types of resolution used in this thesis>

@@

Significant research studies accept the climate change and variability have great impacts on marine ecosystems [@Stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased largely. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean thus presenting regions warming much faster or slower than then the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of the heat content [@nerem; @levitus].

@@

As previously mentioned, upwelling winds in EBUSs are generated by pressure differences between continental thermal lows and oceanic highs. As warming intensifies, pressure gradients between land and sea increases, resulting in a greater intensity of upwelling winds [@bakun2010]. In rare cases, projected upwelling increases may overcome the countervailing effects of upper-ocean warming and stratification to cause regional cooling [@auad]. Research studies reveal declines in ocean oxygen levels as a result of ocean warming and reduced ventilation from stratification and circulation changes (Keeling et al. 2010). Similarly, low oxygen levels are also possible because of shifts in the composition of sinking organic material [@hofmann; @doney).

@@

Extreme thermal events have a range of broad scale definitions, but they can each be classified into two categories, namely, cold-spells [e.g. @gunter; @lirman; @boucek] or heatwaves (e.g. [@gordon; @stott; @perkins]. These events occur rapidly, and unexpectedly and they profoundly impact on species diversity, distributions, growth and survival where they take place. These events are often undetected but studies have found that extreme events have occurred as recently as approximately four decades ago. Jentsch et al. (2007) were the first to discuss this mode of climate change and brought it to focus. Steps have since been taken towards understanding the drivers, impacts and consequences of these events. Subsequently, research has predicted an increase in the occurrence of these extreme events as a result of long-term warming of the planet via the process of global warming (REFERENCE). Extreme events are often associated with toxic algal blooms, shifts in species distribution, and the bleaching of coral reefs [@mccabe; @harley; @schoepf].

>>>>

The duration and intensity of coastal upwelling are known to have a critical role in the phenology of key marine ecosystem processes. Similarly, changes in the characteristics of upwelling are shown to have cause large disturbances to ecosystems at several trophic levels [@wang]. Cold, nutrient rich, low pH, and low oxygen upwelled waters fuels primary production as a consequence of the development of phytoplankton blooms [@ryther; @Cushing; @mcGillicuddy; @oschlies]. In these upwelling systems, the high productivity supports vast and diverse marine populations. Mesoscale and sub-mesoscale oceanic movements are key in managing the exchange and transportation of organic resources and marine organisms from the coastal zone to the open ocean. Most of the world’s major fishing industries are dependent on upwelling, because this process brings nutrient rich waters into the euphotic zone, where it forms the foundation of a localized food web that ecologically differs from surrounding, non-upwelled regions [@bosman; @barshis]. Whilst nutrients play a primary role in the survival of these systems, the cool temperatures of these waters are critical as well. There is an increasingly large body of research that has provided evidence showing that sea surface temperatures (SST) have and are continuing to increase. However, some regions still show a lower or non-significant increase, and in some cases cooling trends are evident (REFERENCES). Upwelling occurring along the Cape south coast is proven to have less of an economic importance in contrast to the west coast. Inshore regions and embayments that are protected from wind action may be subjected to intense episodes of primary production [@graham]. Vander Wouder et al. (2006) found that these sites represent the highest levels of chlorophyll concentration. Biotic systems within the Benguela upwelling system may be negatively influenced by the changes in oxygen levels given their adaption to natural variable conditions [@capone].

@@

Throughout region surrounding the Benguela EBUS, the atmosphere is dominated by anticyclonic high pressure cells with quasi-stationary positions. The South Atlantic Ocean High is situated along the west, drawing cool, dry air onto the west of the subcontinent [@van]. Solar heating during summer may result in the development of low pressure cells, known as heat lows. These cells are absent during winter time [@tyson]. When the anticyclones shift north, during winter months, the cold westerlies have a substantial impact on the weather of the southern tip of the South African subcontinent [@van]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van]. However, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings].

@@

Remotely-sensed satellite SST products are produced and maintained by a number of organisations worldwide. In order to obtain these SST data, advanced technology is launched into space, in order to obtain infra-red light being emitted by the ocean. This irradiance is then taken as a proxy for SST.  Due to the distance at which these SST are being collected, certain coarseness within the datasets are unavoidable. Satellite SST products undergo a range of quality control processes that are well researched and tested [@reynolds], making their use more reliable. However, Smit et al. (2013) have shown that SST data are not particularly reliable for coastal regions and should be used cautiously as they can be prone to both warm and cold bias. Nevertheless, satellite data products are extremely useful when investigating the ocean at broad scale and can provide important SST data.  <will add more here on the different types of resolution used in this thesis>

@@

Significant research studies accept the climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased largely. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean thus presenting regions warming much faster or slower than then the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of the heat content [@nerem; @levitus].

@@

As previously mentioned, upwelling winds in EBUSs are generated by pressure differences between continental thermal lows and oceanic highs. As warming intensifies, pressure gradients between land and sea increases, resulting in a greater intensity of upwelling winds [@bakun2010]. In rare cases, projected upwelling increases may overcome the countervailing effects of upper-ocean warming and stratification to cause regional cooling [@auad]. Research studies reveal declines in ocean oxygen levels as a result of ocean warming and reduced ventilation from stratification and circulation changes (Keeling et al. 2010). Similarly, low oxygen levels are also possible because of shifts in the composition of sinking organic material [@hofmann; @doney].

@@

Extreme thermal events have a range of broad scale definitions, but they can each be classified into two categories, namely, cold-spells [e.g. @gunter; @lirman; @boucek] or heatwaves [e.g. @gordon; @stott; @perkins]. These events occur rapidly, and unexpectedly and they profoundly impact on species diversity, distributions, growth and survival where they take place. These events are often undetected but studies have found that extreme events have occurred as recently as approximately four decades ago. Jentsch et al. (2007) were the first to discuss this mode of climate change and brought it to focus. Steps have since been taken towards understanding the drivers, impacts and consequences of these events. Subsequently, research has predicted an increase in the occurrence of these extreme events as a result of long-term warming of the planet via the process of global warming (REFERENCE). Extreme events are often associated with toxic algal blooms, shifts in species distribution, and the bleaching of coral reefs [@mccabe; @harley; @schoepf].
